---
title: "Ames Project: Part 2"
author: "Andrew Breedlove, Selene Faire"
date: "02-26-2024"
output: html_document
---

# Abstract
In the following project, we've aimed to study the housing market in Ames, Iowa using a dataset which tracked real-estate sales in Ames across a period of 2006-2010 with the
goal of finding the highest performing neighborhoods in the Ames real-estate market, 
determining the most critical factors in real-estate sales prices and then using
those variables to analyze and compare these highest performing neighborhoods towards
some form of ranking across of each across each factor. This would be conducive towards
our overall goal of providing insight into the best neighborhoods in Ames in which to make real-estate investments. 

# Introduction
As many are well aware, housing prices have skyrocketed in our country across the last decade. Looking at per-state housing price growth, the highest rates we've seen across this period is in states such as Idaho where prices have grown +165% (Jones, 2023). Even the states with the least growth in housing prices have seen at least +29.5% in their median housing market list prices, the lowest in this case being Louisiana (Jones, 2023). Iowa is one of the latter states which has seen the least growth over this period, just barely missing the bottom 10 states in median housing price growth and coming at number 11 at +68% (Jones, 2023). If we were to consider the market across the country to be inflated, then Iowa, as one of the least inflated housing markets in the states, could represent a real estate market in its earlier stages of growth which might make for a good market to invest into.

If Iowa as a whole represents an deflated real-estate market, then Ames might represent a valuable city within that to examine in terms of real-estate investments. Ames is ranked number 5 on top 10 metropolitan cities in Iowa with the fasting growing sales price (Redfin), but does not make the list of top 10 most competitive cities in Iowa (Redfin) (meaning that the median number of days that a house stays on the market is not low enough to make the top 10). This means that in Ames, sales prices are growing and the market isn't so overtly competitive that real-estate investments would be difficult to make.

All these things considered, we might ask ourselves the following overarching question: What are the best neighborhoods in Ames, IA for real estate investment? 

# Methods

This section details the methods employed to identify the best neighborhoods for home purchase recommendations using the Ames housing dataset. 

We began by loading the Ames Housing data set. 

```{r}
# Load data set
library("AmesHousing")
ames <- make_ordinal_ames()

dim(ames)
```

We then modified the data set by filtering for “Normal” conditions in the Sale_Condition variable.

Justification: The purpose of this project is to create a common use model to estimate a “typical” sale, therefore, according to the data documentation for this data set (DeCock), it was in our best interest to remove any observations that do not seem typical (such as foreclosures or family sales). Any sale that is not normal does not represent a typical
market transaction and is an outlier in that sense.

```{r}
# Filtered for “Normal” sales conditions
library(tidyverse)
ames <- ames %>%
  filter(Sale_Condition == "Normal")

# Call data dimensions
dim(ames)
```

In order to continue narrowing down our list of neighborhoods for comparison, we then created two (2) new variables to evaluate price per square foot: Lot_Price (lot area price p/sq ft) and Liv_Price (livable area price p/sq ft). 
Using these two new variables, in addition to the Sale Price, we would be able to select the neighborhoods that produce a higher average price in all three categories.
	
```{r}
# Add variable to calculate sale price per square foot, based on lot area
ames$Lot_Price <- ames$Sale_Price / ames$Lot_Area

# Add variable to calculate sale price per square foot, based on livable area
ames$Liv_Price <- ames$Sale_Price / ames$Gr_Liv_Area

dim(ames)
```

We then created a summary data frame (summ.nbhd) for the purpose of grouping the neighborhoods and then summarizing the price categories (total sale, per lot sq ft, per livable sq ft) and count for each neighborhood. 
This would allow us to compare the averages of the 28 individual neighborhoods to the overall averages in the three price categories above.

```{r}
# Summary of price averages by neighborhood
summ.nbhd <- ames %>%
  group_by(Neighborhood) %>%
  summarize(M.Sale_Price = mean(Sale_Price),M.Lot_Price = mean(Lot_Price),M.Liv_Price = mean(Liv_Price),N.Count = n())
summ.nbhd
```

We then filtered the summary data frame to create a list (top.nbhd) which included only neighborhoods with individual means in the three (3) price categories we defined that were above the all-inclusive mean in the same category. We also decided to exclude neighborhoods with a count of less than five sales seeing as we were exploring data that spanned five years and felt that a minimum of five observations per neighborhood was prudent. 
From the list above, we created a list of top neighborhoods for our review which were above average in our three selected price categories. This list produced six neighborhoods for us to delve into: Bloomington Heights, Greens, Northridge, Northridge Heights, Somerset, and Stone Brook.

```{r}
# List of top neighborhoods (above average for Sale Price, Lot Price, and Livable Price)
top.nbhd <- summ.nbhd %>%
  filter(M.Sale_Price >= mean(ames$Sale_Price),M.Lot_Price >= mean(ames$Lot_Price),M.Liv_Price >= mean(ames$Liv_Price),N.Count >= 5)
top.nbhd
```

While our overall purpose in our study is to analyze the differences within our sample of selected neighborhoods, we would still need to be able to visualize the differences between our sample of top neighborhoods and the rest of the neighborhoods in Ames in order to confirm the validity of our selection. To do this, we created one (1) new variable, Top_Nbhd, using the list of top neighborhoods above. This would allow us to distinguish between the neighborhoods we selected and the “others” neighborhoods as a less distinguished set. We named this category ”xOtherx” in order to place the it as alphabetically last in our graphs. 


```{r}
# Add variable to determine if a neighborhood is found in the top.nbhd list, if not categorized as “xOtherx”
ames$Top_Nbhd <- ifelse(ames$Neighborhood %in% top.nbhd$Neighborhood,as.character(ames$Neighborhood),"xOtherx")
	
dim(ames)
```
After establishing the top neighborhoods for real-estate investment in Ames that 
we would be comparing, we needed to establish and test a comparison criteria in the form
of variables we would be using to create a linear regression model. This model would 
allow us to confirm the validity of the variables we would then select and use to conduct our analysese.
Before consulting our sources and attempting to come up with the best numerical and categorical variables for analysis, we decided to run a correlation analysis between sales price and all numerical variables to statistically identify the strongest relationships 
in the numerical variables and provide ourselves with a base model against which to reference the model we would be creating which would include both numerical and categorical
variables. 
The biggest reason for limiting the correlation analysis to numerical variables is that 
we knew there would be at least a few categorical variables that would have a lot of explanatory power in our final model, and we reasoned that we would be better served selecting those variables from our sources rather than from a correlation analysis.


```{r}
# Run correlation analysis of numerical variables to sale price
nbhd_corr <- cor(ames[setdiff(names(ames)[sapply(ames, is.numeric)],"Sale_Price")],ames["Sale_Price"])

# Select top six (7) variables from correlation analysis
which(-nbhd_corr <= sort(-nbhd_corr)[7], arr.ind = TRUE)

# Run linear regression analysis of top variables from correlation to sale price
fit.corr <- lm(Sale_Price ~ Lot_Area + Total_Bsmt_SF+  First_Flr_SF + Gr_Liv_Area + TotRms_AbvGrd + Garage_Cars + Garage_Area, data = ames)

# Model adjusted r squared and graph
summary(fit.corr)$adj.r.squared
plot(fit.corr, which = 1, main = "Correlation")
```

After identifying the most highly correlated numerical variables using the correlation
analysis above, which yielded us an R^2 explanatory score of 0.75 (which is the proportion of variance in our dependent variable Sale Price explained by the independent variables in the linear regression equation), it was time to begin identifying the variables we would use in our new model using some reliable sources.
We decided to select the variables for our linear regression model based on variables
identified from and Opendoor article (Gomez, 2022) which aimed to identify the most
critical factors which influence a home's value. We then mapped these factors to 
corresponding variables in our dataset. These factors and their mappings included:
Type of Home -> [House_Style], Number of Bedrooms -> [Bedroom_AbvGr], Number of Bathrooms 
-> [Bathroom_AbvGr], Square Footage (Lot) -> [Lot_Area], Square Footage (Usable Space) -> 
[Gr_Liv_Area] and Age & Condition / Quality (Exterior Quality) -> [Overall_Qual];

We only needed to newly create one of these variables (Bathroom_AbvGr), which combined
full baths (Full_Bath) and half baths (Half_Bath) into one. Additionally, we originally
included overall condition (Overall_Cond) in the variables we would map to the Age & Condition / Quality factor; However, we ultimately ended up not including this because it had very little explanatory power in our linear regression model summary. It should also be
noted that the linear regression model we built based on the critical factors from our
sources (Gomez, 2022) only included Lot_Area and Gr_Liv_Area as numerical variables
which carried over from the correlation analysis.

```{r}
# Add new variable, addings full baths and half baths
ames$Bathroom_AbvGr <- ames$Full_Bath + ames$Half_Bath

# Run linear regression analysis using opendoor.com's list of features
fit.opendoor <- lm(Sale_Price ~ House_Style + Year_Built + Bedroom_AbvGr + Bathroom_AbvGr + Lot_Area + Gr_Liv_Area + Overall_Qual, data = ames)

# Model adjusted r squared and graph
summary(fit.opendoor)$adj.r.squared
plot(fit.opendoor, which = 1, main = "Opendoor")
```

After putting together our own linear regression model based off of our Opendoor
source for critical factors in home value, which included two new categorical variables
(House_Style & Overall_Qual), we were able to generate a higher R^2 explanatory score
of 0.872 (which is the proportion of variance in our dependent variable Sale Price  explained by the independent variables in the linear regression equation). 

Seeing that we were able to generate a better explanatory set of variables through
consulting our sources, we found it prudent to proceed with our top neighborhood 
comparisons using visualizations of these variables with the ultimate goal of producing
some form of comparison and ranking per factor and variable, allowing anyone interested in investing in these neighborhoods to visualize how they each compare across the most critical 
factors in determining home value.

# Results

## Factors 1: Features (Type of Home, No of Bedrooms and Bathrooms)

Beginning with the factors falling into the features category, we would first visualize
and compare the variables corresponding to Type of Home, Bedrooms and Bathrooms
across the neighborhoods in Ames:

Type of Home (Home Style): First, it is important to note that the house style of approximately 90% of the home observations fall into one of three (3) categories: One-Story (49%), Two-Story (30%) and One-1/2-Story (11%); see Chart 1. That being said, in the majority of neighborhoods two-stories tended to be the highest sold (Graph 1). In addition, neighborhoods Northridge and Northridge Heights came out far ahead in sales price by house style, with Someserset and Stone Brook coming out slightly ahead in the one-story and two-story home styles respectively. 

No of Bedrooms/Bathrooms: The number of bedrooms does not seem to highly impact sale price, when between 2-4 bedrooms (see Graph #2). This is particularly telling seeing as the majority of observations fall within this same range (Chart#2). As for the number of bathrooms, the story seems to change, with more bathrooms leading to a higher sales price, especially for 2-2.5 bathrooms (Graph #3). The majority of observations fall within the 1-2.5 range.
	
```{r}
# Chart 1 - Observations by House Style
ames %>%
  	group_by(House_Style) %>%
  	summarise(Count = n(), Percent = n()/nrow(ames)*100) %>%
  	arrange(desc(Count))

	# Chart 2 - Observations by No. of Bedrooms/Bathrooms
library(janitor)
ames %>%
pivot_wider(id_cols = Bedroom_AbvGr, names_from = Bathroom_AbvGr, values_from = Bathroom_AbvGr, values_fn = list(Bathroom_AbvGr=length), values_fill = 0, names_sort = TRUE) %>%
  	arrange(Bedroom_AbvGr) %>%
  	adorn_totals(c("row","col"))

# Graph 1 - House Style (Boxplot by Neighborhood)
library(scales)
ggplot(ames,aes(Sale_Price, House_Style, fill = Top_Nbhd)) +
  geom_boxplot(data = transform(ames, Top_Nbhd = NULL), fill = "grey85") +
  geom_boxplot() +
  facet_wrap(~Top_Nbhd) +
  labs(title = "Top Neighborhoods: House Style", 
       colour = "Neighborhood") +
  xlab("House Style") + ylab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

# Graph 2 - Bedrooms (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames, aes(Bedroom_AbvGr, Sale_Price, color = Top_Nbhd)) +
  labs(title = "Top Neighborhoods: Bedrooms") +
  xlab("Bedrooms") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

# Graph 3 - Bathrooms (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames, aes(Bathroom_AbvGr, Sale_Price, color = Top_Nbhd)) +
  labs(title = "Top Neighborhoods: Bathrooms (Above Ground)") +
  xlab("Bathrooms") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))
```

## Square Footage (Home Size and Usable Space)

Following features, we would now look into the factors in the Square Footage category,
visualizing and comparing the variables corresponding to Home Size and Usable Space 
across the neighborhoods in Ames:

Home Size (Lot Area): The lot areas for the majority of the selected neighborhoods tended to be smaller and were generally under 25K square feet (see Graph #4). Northridge and Stone Brook did have a few larger lots, but for the most part home size was not hugely significant. 

Usable Space (Liveable Area): Livable area on the other hand, did show a strong relationship with sale price for all neighborhoods (see Graph #5). As far as sales price per square
footage of liveable area, we found Northridge Heights, Stone Brook, and Northridge 
coming out ahead of the other neighborhoods (see Graph #5).

```{r}
	# Graph 4 - Lot Area (Scatter plot by Neighborhood)
ggplot(ames,aes(Lot_Area, Sale_Price, color = Top_Nbhd)) +
  	geom_point(data = transform(ames, Top_Nbhd = NULL), color = "grey85") +
  	geom_point() +
  	facet_wrap(~Top_Nbhd) +
  	labs(title = "Top Neighborhoods: Lot Area") +
  	xlab("Lot Area") + ylab("Sale Price") +
  	scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

# Graph 5 - Livable Area (Scatter plot by Neighborhood) 
  	ggplot(ames,aes(Gr_Liv_Area, Sale_Price, color = Top_Nbhd)) +
  	geom_point(data = transform(ames, Top_Nbhd = NULL), color = "grey85") +
  	geom_point() +
  	facet_wrap(~Top_Nbhd) +
  	labs(title = "Top Neighborhoods: Livable Area") +
  	xlab("Livable Area") + ylab("Sale Price") +
  	scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))
```

## Age & Condition / Quality

Following Square Footage, we would now look into the factors in the Age & Condition category, visualizing and comparing the variables corresponding to such across the neighborhoods in Ames:

Age (Year Built): Generally speaking, we found a strong positive correlation between
year built and sales price, with the most expensive houses sold having been built more
recently. As far as our top neighborhoods, Northridge, Northridge Heights and Stone Brook
came out ahead in this category (see Graph #6).


```{r}
#  Graph 6 - Year Built (Scatter plot by Neighborhood)
ggplot(ames,aes(Year_Built, Sale_Price, color = Top_Nbhd)) +
  geom_point(data = transform(ames, Top_Nbhd = NULL), color = "grey85") +
  geom_point() +
  facet_wrap(~Top_Nbhd) +
  labs(title = "Top Neighborhoods: Year Built", 
       colour = "Neighborhood") +
  xlab("Year Built") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 
```

Condition / Quality (Overall Condition): We found that this variable held little
explanatory significance seeing that the majority of real-estate sales and the highest
sales prices amongst those fell into the "average" overall condition category across
all neighborhoods (see Graph #7). For this reason we removed this variable from our linear regression equation.

Condition / Quality (Overall Quality): We found a strong positive correlation between
overall quality and sales price, and the leading neighborhoods in this category were
once again Northridge Heights, Northridge and Stone Brook (see Graph #8).


```{r}
# Graph 7 - Overall Condition (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames, aes(Overall_Cond, Sale_Price, color = Top_Nbhd)) +
  labs(title = "Top Neighborhoods: Overall Condition") +
  xlab("Quality") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

# Graph 8 - Overall Quality (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames, aes(Overall_Qual, Sale_Price, color = Top_Nbhd)) +
  labs(title = "Top Neighborhoods: Overall Quality") +
  xlab("Quality") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

```

# Discussion

While we found that of our top neighborhoods, Northridge Heights, Northridge, and 
Stone Brook consistently came out on top for most of our critical factors in determing
home value, there is still plenty of nuance when it comes to real-estate investment
and anyone looking to invest in this area would ultimately want to look into which
critical factors are the most important to them and select the top neighborhoods
they would want to look into based on their picks for these factors.

On an factor by factor basis, we can now reflect on the ways our study could have
been improved or have been made more nuanced:

Features (Type of Home, No of Bedrooms and Bathrooms) - While the graph used was able to show the distribution of sale prices amongst the neighborhood by house style, the analysis may have benefitted from a deeper look into the number and proportion of house styles sold by neighborhood. 
For example: Did Northridge or Northridge Heights sell more two-story homes than the other neighborhoods in proportion to how many two-story homes were sold in total during the period covered? Could this be an indicator of not only home but also neighborhood value? 
As for the number of bedrooms and bathrooms, the same could be said: 
Do neighborhoods with homes that have more bedrooms or bathrooms indicate a better investment? 
It would be interesting to explore these questions in a subsequent study. 

Square Footage (Home Size and Usable Space) - Lot area (Home Size) did not seem to indicate a strong relationship with sale price for any of the neighborhoods, and this was 
especially true for the top neighborhoods selected. However, this is not the case for livable area (Usable Space), which seems to be a stronger predictor of sale price across all neighborhoods, including the top neighborhoods selected, with the exception of Bloomington Heights and Greens. A deeper look into these neighborhoods in a subsequent study would be interesting. 


# References

A list of References should include all data sources, R packages, and textbook references, in addition to those that provide background and context for the research problem. It is fine that many references can only be accessed online, especially if you include an actual link. You may choose an appropriate style manual for your field, or use APA style as a default. The style guide itself isn't as important here as empowering an interested reader to find the source themselves. 

## Data Sources

We used The Ames Housing Dataset, which is a well-known dataset in the field of machine learning and data analysis. It contains various features and attributes of residential homes in Ames, Iowa, USA.
This set has 2930 observations and a large number of explanatory variables (23 nominal, 
23 ordinal, 14 discrete, and 20 continuous).

## R Packages

1. "AmesHousing"
2. tidyverse
3. janitor
4. scales

## Citations

1. Jones, Jonathan. “Cities with the Largest Increase in Home Prices over the Last Decade [2023].” Construction Coverage, December 6, 2023. https://constructioncoverage.com/research/cities-with-the-largest-home-price-growth-last-decade#:~:text=Over%20the%20past%20decade%2C%20which,Index%20rose%20by%20just%2032%25. 

2. Redfin. “Iowa Housing Market.” Redfin. Accessed January 26, 2024. https://www.redfin.com/state/Iowa/housing-market. 

3. DeCock, Dean. “Ames Iowa: Alternative to the Boston Housing Data Set.” Jse.amstat.org. Accessed January 26, 2024. https://jse.amstat.org/v19n3/decock/DataDocumentation.txt. 

4. Gomez, Joe. “Sell Your Home the Minute You’re Ready.” Opendoor, June 4, 2022. https://www.opendoor.com/articles/factors-that-influence-home-value. 


