---
title: "Project 2"
author: "Andrew Breedlove, Selene Faire"
date: "2024-02-19"
output: html_document
---

# Abstract

In the following project, we've aimed to study the housing market in Ames, Iowa across
a period of 1950-2010 and select five highest performing neighborhoods in
the market for the purpose of ranking them as the best neighborhoods in which to make real-estate investments. 

## Introduction

As many are well aware, housing prices have skyrocketed in our country across the
last decade. Looking at per-state housing price growth, the highest rates we've seen 
across this period is in states such as Idaho where prices have grown +165% (Jones, 2023).
Even the states with the least growth in housing prices have seen at least +29.5%
in their median housing market list prices, the lowest in this case being Lousiana (Jones, 2023).
Iowa is one of the latter states which has seen the least growth over this period,
just barely missing the bottom 10 states in median housing price growth and coming
in at number 11 at +68% (Jones, 2023). If we were to consider the market across the 
country to be inflated, then Iowa, as one of the least inflated housing markets in the states, could represent a real estate market in its earlier stages of growth which might make for a good market to invest into.

If Iowa as a whole represents an uninflated real-estate market, then Ames might represent
a valuable city within that to examine in terms of real-estate investments. Ames is ranked number 5 on top 10 metropolitan cities in Iowa with the fasting growing sales price (Redfin), but does not make the list of top 10 most competitive cities in Iowa (Redfin) (meaning that the median number of days that a house stays on the market is not low enough to make the top 10). This means that in Ames, sales prices are growing and the market
isn't so overtly competitive that real-estate investments would be difficult to make.

All these things considered, we might ask ourselves the following overarching question:

What are the best neighborhoods in Ames,IA for real estate investment? (Var: Neighborhood, Year_sold, Mo_Sold)


## Methods

We decided to modify our dataset into a working dataset in the following ways:

1. Loading the Ames Housing Data Set

```{r}
library(tidyverse)
library("AmesHousing")

# Create a copy of the cleaned data
ames <- make_ordinal_ames()
```

2. Only houses sold under normal sale condition
Justification: The purpose of this project is to create a common use model to estimate a “typical” sale, therefore, according to the [data documentation](https://jse.amstat.org/v19n3/decock/DataDocumentation.txt) for this data set, it is in our best interest to remove any observations that do not seem typical (such as foreclosures or family sales). With this in mind, the data set has been filtered to only include "Normal" sales conditions. 

```{r}
# Create a copy of the data, filtering for "Normal" sales conditions
ames.norm <- ames %>%
  filter(Sale_Condition == "Normal")

# Call data dimensions
dim(ames.norm)
```

3. Top neighborhoods per total sale price and price per sq ft (lot and livable areas)
Justification: Although many realtors agree that price per square foot doesn't give you a full picture of what a home has to offer, many agree that it is a good starting point to begin an evaluation, as it provides a foundation for the amount of space you're getting compared to the home price. This can be especially useful when looking into neighborhoods, as home values tend to be similar among comparable homes within the same neighborhood. In addition to square footage, a home’s usable space matters when determining its value.

Therefore, to begin our examination of which neighborhoods offer the best opportunity for prospective buyers, we have added two new variables to our data set; Lot Price (lot area price p/sq ft) and Liv Price (livable area price p/sq ft). We will use these two new variables, in addition to the Sale Price, to select the neighborhoods that produce a higher average price in all three categories. 

```{r}
# Add variable to calculate sale price per square foot, based on lot area
ames.norm$Lot_Price <- ames.norm$Sale_Price / ames.norm$Lot_Area

# Add variable to calculate sale price per square foot, based on livable area
ames.norm$Liv_Price <- ames.norm$Sale_Price / ames.norm$Gr_Liv_Area
```

Then we create a summary of neighborhoods by the price categories (total sale, per lot sq ft, per livable sq ft) and count. We use the summary to compare the neighborhood averages to the data set averages in the three price categories and, based on neighborhoods that are above average in each price category, create a list of the top neighborhoods for our review. 

We are also excluding two neighborhoods (Landmark and Green Hills) with less than five sales since we are exploring data that spans five years and feel that a minimum of five observations per neighborhood is prudent. 

```{r}
# Summary of Sale Price, Lot (Area) Price, Livable (Area) Price averages by neighorhood
summ.nbhd <- ames.norm %>%
  group_by(Neighborhood) %>%
  summarize(M.Sale_Price = mean(Sale_Price),M.Lot_Price = mean(Lot_Price),M.Liv_Price = mean(Liv_Price),N.Count = n())

summ.nbhd

# Summary of number of observations by neighborhood
ames.norm %>%
  group_by(Neighborhood) %>%
  summarize(n_count = n()) %>%
  arrange(n_count)

# List of top neighborhoods (above average for Sale Price, Lot Price, and Livable Price)
top.nbhd <- summ.nbhd %>%
  filter(M.Sale_Price >= mean(ames.norm$Sale_Price),M.Lot_Price >= mean(ames.norm$Lot_Price),M.Liv_Price >= mean(ames.norm$Liv_Price),N.Count >= 5)

top.nbhd
dim(top.nbhd)
```

Finally, we create a new data set based on this list. We have also created  two other data sets to compare to our top neighborhoods. The first transforms the neighborhood variable for neighborhoods not listed in our top neighborhoods to the name "Other". The second data set is then based on these neighborhoods. 

```{r}
# Create a copy of the data, filtering for "Top" neighborhoods
 ames.nbhd <- ames.norm %>%
  filter(Neighborhood %in% top.nbhd$Neighborhood)

dim(ames.nbhd)

# Create a copy of the data, transforming neighborhood names not in our list to "xOtherx"
ames.transf <- transform(ames.norm, Neighborhood = factor(replace(as.character(Neighborhood),list = !Neighborhood %in% top.nbhd$Neighborhood,values = "xOtherx")))

dim(ames.transf)

# Create a copy of the data, filtering for "xOtherx" neighborhoods
ames.other <- ames.transf %>%
  filter(Neighborhood == "xOtherx")

dim(ames.other)
```

This list produced six neighborhoods for us to delve into: Bloomington Heights, Greens, Northridge, Northridge Heights, Somerset, and Stone Brook. These neighborhoods together have an average sale price of $270K, which is 53% higher than the average sale price of $176K for all neighborhoods, which makes them prime candidates for our analysis.

```{r}
# Summary of number of observations by "top" neighborhood
ames.nbhd %>%
  group_by(Neighborhood) %>%
  summarize(n_count = n()) %>%
  arrange(desc(n_count))

# Statistical summaries of Sale Price for all three datasets
summary(ames.nbhd$Sale_Price)
summary(ames.transf$Sale_Price)
summary(ames.other$Sale_Price)

# Histograms of Sale Price for all three datasets
par(mfrow=c(1,3))
hist(ames.norm$Sale_Price, main = "All Neighborhoods", xlab = "Sale Price")
hist(ames.nbhd$Sale_Price, main = "Top Neighborhoods", xlab = "Sale Price")
hist(ames.other$Sale_Price, main = "Other Neighborhoods", xlab = "Sale Price")

#XXXXX ***** XXXXX NOT SURE IF USEABLE

library(scales)

ggplot() +
  geom_histogram(data = ames.other, aes(Sale_Price), fill = "gray") +
  geom_histogram(data = ames.nbhd, aes(Sale_Price, color = Neighborhood)) +
  labs(title = "Other v Top Neighborhoods: Sale Price") +
  xlab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

ggplot(ames.nbhd,aes(Sale_Price,Neighborhood)) +
  geom_point(aes(color = Neighborhood)) +
  labs(title = "Top Neighborhoods: Sale Price") +
  xlab("Sale Price") + ylab("Neighoborhood") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

ggplot(ames.transf,aes(Sale_Price)) +
  geom_histogram(aes(color = Neighborhood)) +
  labs(title = "Top Neighborhoods: Sale Price") +
  xlab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

ggplot(ames.nbhd,aes(Sale_Price)) +
  geom_histogram(aes(color = Neighborhood)) +
  labs(title = "Top Neighborhoods: Sale Price") +
  xlab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 
```

XXXXX ***** XXXXX NOT SURE IF USEABLE

4. Critical factors that influence a home's value
Justification: Now that we have selected our top neighborhoods, we return to [Opendoor.com](https://www.opendoor.com/articles/factors-that-influence-home-value) to evaluate the factors that are considered critical to influencing a home's value. The first factor is neighborhood comps. Opendoor provides advice to finding comps by suggesting a look at recency, feature similarities, condition & quality, distance and location. However, because our data set does not offer sufficient information on factors for evaluating distance (HOA rules, school districts, etc.) we will not be including this "critical factor" in our analysis. ***Location***

4a. Neighborhood Comps: Recency
We begin our analysis with recency. Given the limits of our data set, we have decided to evaluate the year sold to determine if there are any patterns to the relationship between the year sold and the sale price. 

For this step, we decided to find the aggregated mean of the Sale Price by Neighborhood and Year Sold and then to plot the results by each top neighborhood. We have also included a graph to compare the group to the rest or "Other" neighborhoods. As can be seen, the majority of the neighborhoods show a dip in sales as they headed into 2010. This makes sense as that aligns with the housing market collapse of 2008. This can be seen in almost all of the neighborhoods, including the "Other" neighborhood grouping, with the exception of Stone Brook, which bears keeping an eye on. 

```{r}
# Summary of Sale Price averages by neighorhood and year sold
nbhd.yr.sold <- ames.transf %>%
  select(Sale_Price,Neighborhood,Year_Sold) %>%
  group_by(Neighborhood,Year_Sold) %>%
  summarise(M.Sale_Price = mean(Sale_Price))

# Graph of Sale Price average by year sold by neighborhood
ggplot(nbhd.yr.sold,aes(Year_Sold, M.Sale_Price, color=Neighborhood)) +
  geom_point() +
  facet_wrap(~Neighborhood) +
   labs(title = "Top Neighborhoods: Recency (Year Sold)", 
       colour = "Neighborhood") +
  xlab("Year Sold") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

```

4b. Neighborhood Comps: Feature Similarities

For feature similarities, we also use Opendoor's list as a guide, which lists type of home, year built, number of bedrooms, bathrooms and square footage. We will use House Style for type of home, limit our the number of bedrooms and bathrooms to above ground, and use both lot and livable area for square footage. 

We start by creating a new variable for the number of bathrooms by combining full baths and half baths. We decided not to include the bathrooms in the basement in order to maintain consistency with the number of bedrooms and livable area, which are both referring to areas that are above ground. 

At this point, we felt that this would be a good point to start building our model using the features selected. Using the linear regression function, we were able to calculate an 83% adjusted R squared, indicating a high model accuracy. 

However, we were also curious to see if an alternate model based on the highest correlated variables might show greater accuracy. Surprisingly, the model  also showed an 83% adjusted R squared with about a half percent difference. 
 
```{r}
# Add new variable, addings full baths and half baths
ames.nbhd$Bathroom_AbvGr <- ames.nbhd$Full_Bath + (0.5 * ames.nbhd$Half_Bath)
ames.transf$Bathroom_AbvGr <- ames.transf$Full_Bath + (0.5 * ames.transf$Half_Bath)

# Create linear regression model using opendoor.com's list of features
fit.opendoor <- lm(Sale_Price ~ House_Style + Year_Built + Bedroom_AbvGr + Bathroom_AbvGr + Lot_Area + Gr_Liv_Area, data = ames.nbhd)

summary(fit.opendoor)
# Model adjusted r squared and graph
summary(fit.opendoor)$adj.r.squared
plot(fit.opendoor, which = 1, main = "Opendoor")

# Correlation summary of numerical variables to Sale Price
nbhd_corr <- cor(ames.nbhd[setdiff(names(ames.nbhd)[sapply(ames.nbhd, is.numeric)],"Sale_Price")],ames.nbhd["Sale_Price"])

# Top six variables correlating to Sale Price
which(-nbhd_corr <= sort(-nbhd_corr)[6], arr.ind = TRUE)

# Create linear regression model using correlation result list
fit.corr <- lm(Sale_Price ~ Gr_Liv_Area + First_Flr_SF + Total_Bsmt_SF+ Lot_Area + TotRms_AbvGrd + Garage_Area, data = ames.nbhd)

# Model adjusted r squared and graph
summary(fit.corr)$adj.r.squared
plot(fit.opendoor, which = 1, main = "Correlation")

# House Style (Boxplot by Neighborhood)
ggplot(ames.transf,aes(Sale_Price, House_Style, fill = Neighborhood)) +
  geom_boxplot(data = transform(ames.transf, Neighborhood = NULL), fill = "grey85") +
  geom_boxplot() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: House Style", 
       colour = "Neighborhood") +
  xlab("House Style") + ylab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

# Year Built (Scatter plot by Neighborhood)
ggplot(ames.transf,aes(Year_Built, Sale_Price, color = Neighborhood)) +
  geom_point(data = transform(ames.transf, Neighborhood = NULL), color = "grey85") +
  geom_point() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: Year Built", 
       colour = "Neighborhood") +
  xlab("Year Built") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

# Bedrooms (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames.transf, aes(Bedroom_AbvGr, Sale_Price, color = Neighborhood)) +
  labs(title = "Top Neighborhoods: Bedrooms") +
  xlab("Bedrooms") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

# Bathrooms (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames.transf, aes(Bathroom_AbvGr, Sale_Price, color = Neighborhood)) +
  labs(title = "Top Neighborhoods: Bathrooms (Above Ground)") +
  xlab("Bathrooms") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

# Lot Area (Scatter plot by Neighborhood)
ggplot(ames.transf,aes(Lot_Area, Sale_Price, color = Neighborhood)) +
  geom_point(data = transform(ames.transf, Neighborhood = NULL), color = "grey85") +
  geom_point() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: Lot Area") +
  xlab("Lot Area") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

# Livable Area (Scatter plot by Neighborhood) 
  ggplot(ames.transf,aes(Gr_Liv_Area, Sale_Price, color = Neighborhood)) +
  geom_point(data = transform(ames.transf, Neighborhood = NULL), color = "grey85") +
  geom_point() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: Livable Area") +
  xlab("Livable Area") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))
```
4c. Neighborhood Comps: Distance
****ADD NARRATIVE TO TIE TO RESULTS (I.E. HOA FOR TOP PICK NEIGHBORHOODS)??? ****

4d. Neighborhoods Comps: Condition & Quality

Opendoor stresses that there is quite a bit of value in knowing that repairs will not
be necessary for a while after purchasing a home. This depends on the condition of 
the property when it is bought, as well as the materials and finish which make up 
the initial quality of the home. The better the materials and finish, the longer a 
buyer will have before they need to pay for repairs, adding value to the sales price
of a home. 
For this category, we'll be looking at the Overall_Cond variable for condition and
the Overall_Qual variable for quality:

```{r}
# Overall Condition (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames.transf, aes(Overall_Cond, Sale_Price, color = Neighborhood)) +
  labs(title = "Top Neighborhoods: Overall Condition") +
  xlab("Quality") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

# Overall Quality (Scatter plot by Neighborhood)
ggplot() +
  geom_point(data = ames.transf, aes(Overall_Qual, Sale_Price, color = Neighborhood)) +
  labs(title = "Top Neighborhoods: Overall Quality") +
  xlab("Quality") + ylab("Sale Price") +
  scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

```
As we can see from the above, overall condition does not seem consistently trend
with the sales price of a home seeing as the houses with the highest sales prices
fell into the "average" condition category. This includes pretty much all of the 
houses in our top neighborhoods.
Overall quality however, seems to have a strong positive correlation with sales price
and we can see from the graph that the houses in the neighborhoods we selected as
the highest performing also have the highest overall quality of material and finish.
Seeing as such, we'll go ahead and add Overall_Qual to our Opendoor influenced linear regression model and see how much it improves our R^2 value:

```{r}
# Create linear regression model using opendoor.com's list of features
fit.opendoor <- lm(Sale_Price ~ House_Style + Year_Built + Bedroom_AbvGr + Bathroom_AbvGr + Lot_Area + Gr_Liv_Area + Overall_Qual, data = ames.nbhd)

summary(fit.opendoor)
# Model adjusted r squared and graph
summary(fit.opendoor)$adj.r.squared
plot(fit.opendoor, which = 1, main = "Opendoor")
```
As we can see above, adding Overall_Quality to our regression model improved our
R^2 value by nearly 7%, making it a significantly correlated variable we'll be keeping
in our final model.

4e. Neighborhood Comps: Location
For location or placement as opendoor.com calls it, we have decided to look at a couple more variables, namely MS Zoning, Land Countour, Land Slope, and Condition 1 or proximity to various conditions such as arterial or feeder streets, railroads, or off-site features such as parks, greenbelts, etc.

```{r}
# MS Zoning
  ggplot(ames.transf,aes(Sale_Price, MS_Zoning, fill = Neighborhood)) +
  geom_boxplot(data = transform(ames.transf, Neighborhood = NULL), fill = "grey85") +
  geom_boxplot() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: Zoning", 
       colour = "Neighborhood") +
  xlab("Zoning") + ylab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

# Land Contour
  ggplot(ames.transf,aes(Sale_Price, Land_Contour, fill = Neighborhood)) +
  geom_boxplot(data = transform(ames.transf, Neighborhood = NULL), fill = "grey85") +
  geom_boxplot() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: Land Contour", 
       colour = "Neighborhood") +
  xlab("Land Contour") + ylab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

# Land Slope
  ggplot(ames.transf,aes(Sale_Price, Land_Slope, fill = Neighborhood)) +
  geom_boxplot(data = transform(ames.transf, Neighborhood = NULL), fill = "grey85") +
  geom_boxplot() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: Land Slope", 
       colour = "Neighborhood") +
  xlab("Land Slope") + ylab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

# Condition 1 (Proximity to various conditions)
  ggplot(ames.transf,aes(Sale_Price, Condition_1, fill = Neighborhood)) +
  geom_boxplot(data = transform(ames.transf, Neighborhood = NULL), fill = "grey85") +
  geom_boxplot() +
  facet_wrap(~Neighborhood) +
  labs(title = "Top Neighborhoods: Proximity", 
       colour = "Neighborhood") +
  xlab("Proximity") + ylab("Sale Price") +
  scale_x_continuous(labels = label_number(suffix = "K", scale = 1e-3)) 

# **** ADD ANY TO MODEL? ONLY LAND CONTOUR INCREASES TO 84.03%. MATERIAL? MAY NEED MAP ****

```





## Results


## Discussion



## Citations

1. Jones, Jonathan. “Cities with the Largest Increase in Home Prices over the Last Decade [2023].” Construction Coverage, December 6, 2023. https://constructioncoverage.com/research/cities-with-the-largest-home-price-growth-last-decade#:~:text=Over%20the%20past%20decade%2C%20which,Index%20rose%20by%20just%2032%25. 

2. Redfin. “Iowa Housing Market.” Redfin. Accessed January 26, 2024. https://www.redfin.com/state/Iowa/housing-market. 
